Determinism, Complexity, and Predictability in Computer Performance

### 2014-03-22 model.tex
making notation consistent in model.tex

A simple and yet powerful way to capture and exploit the structure of
data is to fit a hyperplane to the dataset and then use that it to
make predictions.  The roots of this approach date back to Yule's 1927
invention of the autoregressive schema~\cite{weigend93}, which
forecasts the next time step through a weighted average of past
observations: $$p_i = \sum_{j=1}^{i-1} a_j x_j$$ The weighting
coefficients $a_i$ are computed using a polynomial of degree $p$ that
is fit to the past $p+1$ values of $x_i$.  To account for noise in the
data, one can add a so-called ``moving average'' term to the model; to
remove nonstationarities, one can detrend the data using a
differencing operation.  A strategy that incorporates all three of
these features is called a \emph{nonseasonal ARIMA model}.  If
evidence of periodic structure is present in the data, a
\emph{seasonal ARIMA model}, which adds a sampling operation that
filters out periodicities, can be a good choice.

There is a vast amount of theory and literature regarding the
construction and use of models of this type; we refer the reader to
\cite{davislinearts} for an in-depth exploration.  For the purposes of
this paper, where the goal is to explore the relationship between
predictability and complexity across a broad array of forecast
strategies, seasonal ARIMA models are a good exemplar of the class of
linear predictors.  Fitting such a model to a dataset involves
choosing values for the various free parameters in the autoregressive,
detrending, moving average, and filtering terms---e.g., the $p$ in the
equation at the beginning of this section.  We employ the automated
fitting techniques described in~\cite{autoARIMA} to accomplish this.
This procedure uses sophisticated methods---KPSS unit-root
tests~\cite{KPSSunit}, a customization of the Canova-Hansen
test~\cite{Canova1995}, and the Akaike information
criterion~\cite{akaike1974}, conditioned on the maximized likelihood
of the model fitted to the detrended data---to select good values for
the free parameters of the ARIMA model.

@@

      \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colLMAForecast.png}
    \caption{\col \\ LMA}
    \label{fig:colLMA}
  \end{subfigure}
      \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/gccLMAForecast.png}
    \caption{\gcc \\ LMA}
    \label{fig:gccLMA}
  \end{subfigure}  
        \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/svdfiveLMAForecast.png}
    \caption{\svdfive \\ LMA}
    \label{fig:gccLMA}
  \end{subfigure}
  %\begin{subfigure}{0.5\textwidth}
  %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
   \caption{Predicted ($p_j$) versus true values ($c_j$) for
     forecasts of \col, \gcc, and \svdfive.  On this type of plot, a
     perfect prediction lies exactly on the diagonal.
%
{\color{red} Joshua, what would you think of adding a row of
  random-walk plots at the top?  The letters in the subcaptions---(a)
  (b) etc.---feel clunky here.  Finally, we should eventually change
  the label of the vertical axes on these plots to be whatever symbol
  we choose for the prediction.}
%
}
\label{fig:forecast-example}
  %\end{subfigure}
\end{figure*} 
In these images, the vertical axis is the prediction $p_j$ and the
horizontal axis is the true continuation $c_j$.  On such a plot, a
perfect prediction would lie on the diagonal.  LMA, for instance,
generates a very accurate prediction of the \col data, while ARIMA
does not.  [[We should comment on what a noise ball means on such a
    plot, as well as what a horizontal line means.]]

As a numerical measure of prediction accurary, we calculate the mean
absolute scaled error ($MASE$) between the true and predicted signals:
% 
$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|c_j-x_j|
}{\frac{k}{n-1}\sum^n_{i=2}|x_{i}-x_{i-1}|}$$
% 
This error metric was introduced in \cite{MASE} as a ``generally
applicable measurement of forecast accuracy without the problems seen
in the other measurements."  $MASE$ is a normalized measure; the
scaling term in the denominator
% 
% of the MASE value
% $$\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|$$ 
% 
is the average in-sample forecast error for a random-walk prediction
over a signal of length $k$.  That is, $MASE<1$ means that the
prediction error in question was, on the average, smaller than the
error of a random-walk forecast on the same data.  Analogously,
$MASE>1$ means that the corresponding prediction method did
\emph{worse}, on average, than the random-walk method.  We chose this
error metric because it allows for fair comparison across varying
methods, prediction horizons, and signal scales.

@@

 \svdsix     & $ 0.739\pm 0.068 $ & $ 1.438\pm 0.0610$ & $2.197\pm0.083$&  $0.748 \pm 0.011$ \\
  \hline
  \end{tabular}
  \end{center}
 \label{default}
 \caption{MASE scores and weighted permutation entropies for all eight
   processes studied in this paper}
 \label{tab:error}
  \end{table*}%
Comparing these with the geometry of the plots in
Figure~\ref{fig:forecast-example}, one can see some obvious
correspondences.  The average LMA $MASE$ score for the \col signals
was $0.050$, for instance, while ARIMA scored much worse ($0.599$).
That is, LMA performed roughly 20 times better on \col signals than a
random-walk predictor, while ARIMA only outperformed random walk by a
factor of 1.7 [[is this right?  there was a ``2'' here, but 1/0.599 is
    closer to 1.7]].  This is in accordance with the visual appearance
of the corresponding images in Figure~\ref{fig:forecast-example}.
While the correspondence between $MASE$ score and the visual
appearance of a $p$ versus $c$ plot is clear cut in this case, that
is not always so.  The plots of LMA predictions of \col and \svdfive
both lie near the diagonal, for instance, but the corresponding $MASE$
scores were very different: $0.050$ for \col and $0.718$ for \svdfive
(resp., 20 and 1.4 times better than random-walk forecasts of the same
signals).  This is a result of the normalization in the $MASE$ score.
Recall from Section~\ref{sec:simple} that the random-walk method
performs especially poorly on signals that oscillate rapidly.  \col
fits this description, so the random-walk error on this signal---the
denominator of the three $MASE$ scores in the first row of the
Table---is pathologically high, which skews those scores down.
Random-walk prediction is very effective for the\svdfive signal, on
the other hand, so the denominator is small and the $MASE$ scores are
skewes upwards.  Normalization is a two-edged sword here.  It
facilitates comparisons across signals of different types and lengths,
but its particular pathologies must be taken into account, as
discussed at more length in Section~\ref{sec:results}.

>>>>

A simple and yet powerful way to capture and exploit the structure of
data is to fit a hyperplane to the dataset and then use that it to
make predictions.  The roots of this approach date back to Yule's 1927
invention of the autoregressive schema~\cite{weigend93}, which
forecasts the next time step through a weighted average of past
observations: $$p_i = \sum_{j=1}^{i-1} a_j x_j$$ The weighting
coefficients $a_i$ are computed using a polynomial of degree $l$ that
is fit to the past $l+1$ values of $x_i$.  To account for noise in the
data, one can add a so-called ``moving average'' term to the model; to
remove nonstationarities, one can detrend the data using a
differencing operation.  A strategy that incorporates all three of
these features is called a \emph{nonseasonal ARIMA model}.  If
evidence of periodic structure is present in the data, a
\emph{seasonal ARIMA model}, which adds a sampling operation that
filters out periodicities, can be a good choice.

There is a vast amount of theory and literature regarding the
construction and use of models of this type; we refer the reader to
\cite{davislinearts} for an in-depth exploration.  For the purposes of
this paper, where the goal is to explore the relationship between
predictability and complexity across a broad array of forecast
strategies, seasonal ARIMA models are a good exemplar of the class of
linear predictors.  Fitting such a model to a dataset involves
choosing values for the various free parameters in the autoregressive,
detrending, moving average, and filtering terms---e.g., the $l$ in the
equation at the beginning of this section.  We employ the automated
fitting techniques described in~\cite{autoARIMA} to accomplish this.
This procedure uses sophisticated methods---KPSS unit-root
tests~\cite{KPSSunit}, a customization of the Canova-Hansen
test~\cite{Canova1995}, and the Akaike information
criterion~\cite{akaike1974}, conditioned on the maximized likelihood
of the model fitted to the detrended data---to select good values for
the free parameters of the ARIMA model.

@@

      \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colLMAForecast.png}
    \caption{\col \\ LMA}
    \label{fig:colLMA}
  \end{subfigure}
      \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/gccLMAForecast.png}
    \caption{\gcc \\ LMA}
    \label{fig:gccLMA}
  \end{subfigure}  
        \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/svdfiveLMAForecast.png}
    \caption{\svdfive \\ LMA}
    \label{fig:gccLMA}
  \end{subfigure}
  %\begin{subfigure}{0.5\textwidth}
  %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
   \caption{Predicted ($p_j$) versus true values ($c_j$) for
     forecasts of \col, \gcc, and \svdfive.  On this type of plot, a
     perfect prediction lies exactly on the diagonal.
%
{\color{red} Joshua, what would you think of adding a row of
  random-walk plots at the top?  Also, the letters in the
  subcaptions---(a) (b) etc.---feel clunky here.  What would you think
  of removing them?}
%
}
\label{fig:forecast-example}
  %\end{subfigure}
\end{figure*} 
In these images, the vertical axis is the prediction $p_j$ and the
horizontal axis is the true continuation $c_j$.  On such a plot, a
perfect prediction would lie on the diagonal.  LMA, for instance,
generates a very accurate prediction of the \col data, while ARIMA
does not.  {\color{red}[[We should comment on what a noise ball means
      on such a plot, as well as what a horizontal line means.]]}

As a numerical measure of prediction accurary, we calculate the mean
absolute scaled error ($MASE$) between the true and predicted signals:
% 
$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|p_j-x_j|
}{\frac{k}{n-1}\sum^n_{i=2}|x_{i}-x_{i-1}|}$$
% 
{\color{red} That equation had $c_j$ instead of $p_j$ in the numerator.
I changed it.  Let me know if that's wrong.}
This error metric was introduced in \cite{MASE} as a ``generally
applicable measurement of forecast accuracy without the problems seen
in the other measurements."  $MASE$ is a normalized measure; the
scaling term in the denominator
% 
% of the MASE value
% $$\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|$$ 
% 
is the average in-sample forecast error for a random-walk prediction
over a signal of length $k$.  That is, $MASE<1$ means that the
prediction error in question was, on the average, smaller than the
error of a random-walk forecast on the same data.  Analogously,
$MASE>1$ means that the corresponding prediction method did
\emph{worse}, on average, than the random-walk method.  We chose this
error metric because it allows for fair comparison across varying
methods, prediction horizons, and signal scales.

@@

 \svdsix     & $ 0.739\pm 0.068 $ & $ 1.438\pm 0.0610$ & $2.197\pm0.083$&  $0.748 \pm 0.011$ \\
  \hline
  \end{tabular}
  \end{center}
 \label{default}
 \caption{MASE scores and weighted permutation entropies for all eight
   processes studied in this paper}
 \label{tab:error}
  \end{table*}%
Comparing these with the geometry of the plots in
Figure~\ref{fig:forecast-example}, one can see some obvious
correspondences.  The average LMA $MASE$ score for the \col signals
was $0.050$, for instance, while ARIMA scored much worse ($0.599$).
That is, LMA performed roughly 20 times better on \col signals than a
random-walk predictor, while ARIMA only outperformed random walk by a
factor of 1.7 {\color{red}[[is this right?  there was a ``2'' here,
      but 1/0.599 is closer to 1.7]]}.  This is in accordance with the
visual appearance of the corresponding images in
Figure~\ref{fig:forecast-example}.  While the correspondence between
$MASE$ score and the visual appearance of a $p$ versus $c$ plot is
clear cut in this case, that is not always so.  The plots of LMA
predictions of \col and \svdfive both lie near the diagonal, for
instance, but the corresponding $MASE$ scores were very different:
$0.050$ for \col and $0.718$ for \svdfive (resp., 20 and 1.4 times
better than random-walk forecasts of the same signals).  This is a
result of the normalization in the $MASE$ score.  Recall from
Section~\ref{sec:simple} that the random-walk method performs
especially poorly on signals that oscillate rapidly.  \col fits this
description, so the random-walk error on this signal---the denominator
of the three $MASE$ scores in the first row of the Table---is
pathologically high, which skews those scores down.  Random-walk
prediction is very effective for the \svdfive signal, on the other
hand, so the denominator is small and the $MASE$ scores are skewes
upwards.  Normalization is, as is often the case, a two-edged sword.
Here, it facilitates comparisons across signals of different types and
lengths, but its particular pathologies must be taken into account, as
discussed at more length in Section~\ref{sec:results}.
